\section{Related Work}

\paragraph{Tool Use and External Computation in LLMs.}
A growing body of work augments language models with external tools \citep{tool-learning-survey-2024}, spanning several functional categories.
Self-supervised methods learn tool invocation from loss signals \citep{toolformer-2023,toolace-2025}, token-level methods embed tools in the vocabulary \citep{toolkengpt-2023}, task-level decomposition routes subtasks to specialized modules \citep{art-2023,chameleon-2023}, and RL-based methods learn routing from reward \citep{retool-2025}.
All focus on \emph{learning to use tools}; we ask a prior question: \emph{what signal drives the routing decision?}
Concurrent work examines \emph{when} to invoke tools: tool-call decisions correlate with uncertainty~\citep{tool-overuse-illusion-2026}, are encoded in hidden states~\citep{when2tool-2026}, impose a ``tool-use tax''~\citep{tool-use-tax-2026}, or follow necessity-utility-affordability logic~\citep{to-call-or-not-2026}. Our decomposition yields a concrete answer: 78.9\% of non-tie routing labels trace to catastrophic-failure prevention.

\paragraph{Adaptive Computation and Routing.}
Adaptive routing traces back to learned halting policies \citep{graves2016act}. Mixture-of-Recursions \citep{mixture-of-recursions-2025} learns token-level recursion depths in the LLM era; concurrent work further diversifies routing granularity: CogRouter \citep{cogrouter-2026} adapts per-step depth, ARES \citep{ares-2026} selects effort levels, and MTRouter \citep{mtrouter-2026} addresses multi-turn settings; in death-dominated environments, the task-type-level binary signal is structurally inaccessible to these step-level approaches (\S\ref{sec:router-diagnostic}).
RouteLLM \citep{routellm-2024} learns routers from preference data to select between strong and weak models; cost-optimization routing (FrugalGPT~\citep{frugalgpt-2023}; AutoMix~\citep{automix-2024}) cascades queries across model tiers. These methods optimize \emph{which} model to invoke; our framework diagnoses \emph{what} drives the routing decision.
Concurrently, \citet{budget-aware-routing-2026} train episode-level routing policies under sparse terminal rewards, the closest setting to ours.
Unlike token-level adaptive computation (Mixture-of-Depths~\citep{mixture-of-depths-2024}; speculative decoding~\citep{leviathan2023speculative}; early exit~\citep{xin2020deebert,zhou2020bert}) and Mixture-of-Experts architectures~\citep{switch-transformer-2022,gshard-2021}, our framework operates at episode-level granularity, diagnosing aggregate routing signal composition rather than optimizing per-token compute allocation. Counterfactual data augmentation~\citep{kaushik2020learning} edits inputs to flip labels; our framework executes full counterfactual \emph{episodes} to decompose routing signal at the environment level.

\paragraph{Counterfactual Evaluation in RL.}
Prior work generates counterfactual explanations by perturbing individual actions \citep{olson2021counterfactual, gajcin2023counterfactual, amitai2024explaining}; recent and concurrent extensions target LLM agents~\citep{pona2025abstract,axis-2026,cso-2026}. We instead execute complete episodes under two strategies to decompose routing signal \emph{before} training.

\paragraph{LLM Agent Environments and Reasoning.}
ScienceWorld \citep{scienceworld-2022} serves as our primary testbed; ALFWorld \citep{alfworld-2021} provides a complementary environment; AgentBench \citep{agentbench-2024} offers systematic evaluation. ReAct \citep{react-2023} serves as our internal baseline; Reflexion \citep{reflexion-2023} extends this with verbal self-reflection; LATS \citep{lats-2024} adds tree search. The concurrent SALT~\citep{salt-2026} assigns step-level credit on ALFWorld, complementing our episode-level decomposition.

\paragraph{Safe RL and Causal Grounding.}
Death avoidance connects to constrained MDPs \citep{altman1999constrained}, risk-sensitive RL \citep{mihatsch2002risk}, safe RL \citep{garcia2015comprehensive,achiam2017cpo,moldovan2012safe}, and shielding \citep{alshiekh2018safe}; CVaR-based methods~\citep{chow2015risk,tamar2015optimizing} optimize against worst-case outcomes; our framework instead diagnoses which signals derive from catastrophic risk. Classical methods assume known dynamics or differentiable constraints, unavailable for LLM agents. \citet{ruan2024failures} identify catastrophic actions as a failure category; SafeRoute \citep{saferoute-2025} and Agent-SafetyBench \citep{agent-safetybench-2024} address model selection and systematic evaluation for agent safety. Our pure-strategy comparison is a simplified instance of the potential outcomes framework~\citep{rubin1974estimating}, eliminating confounding without structural causal models~\citep{pearl2009causality}. \citet{bonagiri2025check} show selective quitting yields safety gains; our framework explains \emph{why}---death avoidance dominates routing signal. The grounding failures underlying death events connect to hallucination detection~\citep{manakul2023selfcheckgpt,kadavath2022language}, where incorrect state beliefs produce catastrophic outcomes.